Compositional Certified Resource Bounds

### 2014-07-03 paper/main.tex
add comparison results
explaining the experiments

\begin{figure*}
\centering
\begin{tabular}{r|c|c|c|c}
%
&
\begin{lstlisting}
while (y > x) {
  x++; tick(1); }
while (x > 2) {
  x -= 3; tick(1); }
\end{lstlisting}
&
\begin{lstlisting}
while (i > 100) {
  i--; tick(1) }
i += k+50;
while (i >= 0) {
  i--; tick(1); }
\end{lstlisting}
&
\begin{lstlisting}
while (x > 0) {
  x--;
  t=x, x=y, y=t;
  tick(1);
}
\end{lstlisting}
&
\begin{lstlisting}
assert(y >= 0);
while (x>y) {
  x -= y+1;
  for (z=y; z > 0; z--)
    tick(1);
  tick(1);
}
\end{lstlisting}
\\
%
\hline
& t08a (slightly modified t08) & t19 & t30 & t15
%
\\
\hline
LOOPUS &
{\small $\max(0, x{-}2) {+} 2\max(0, y {-} x)$} &
{\small $\max(0, \! i{-}10^2) \!+ \max(0, \! k {+} i {+} 51)$} &
{\small % $1 + \max(x, y)$ (???)
---
} &
---
\\
%
%
%
%
Rank &
{\small $2 {+} y {-} x (???)$ } &
{\small $54+k+i$ } &
--- &
{\small $2+2x-y$ }
\\
%
%
AAA 
  & {\small $1.33|[x,y]| {+} 0.33 |[0,x]|$ }
  & $50 {+} |\inter {-1} i| {+} |\inter 0 k|$ 
  & $|\inter 0 x| {+} |\inter 0 y|$
  & $|\inter 0 x|$
\end{tabular}
\caption{ Comparison of resource bounds derived by different tools on
  several examples with linear bounds.  AAA stands for our automatic
  amortized analysis for Clight.  The output of Rank has been manually
  simplified to fit the table.  }
\label{fig:compar}
\end{figure*}

@@

\begin{figure*}
\centering
\begin{tabular}{r|c|c|c|c|c}
& KoAT & Rank & LOOPUS & SPEED & AAA \\
\hline
\#bounds &
\\
\#lin.\ bounds &
\\
\#best bounds &

\end{tabular}
\caption{Comparison of our automatic amortized analysis with other
  automatic tools.  We have not been able to obtain a version of
  SPEED~\cite{GulwaniMC09} and just use the bounds that have been
  reported by the authors. Similarly, KoAT~\cite{BrockschmidtEFFG14}
  does currently not work on C programs and we use the bounds that
  have been reported in the author's experimental evaluation.}
\label{fig:compar}
\end{figure*}

@@

File &
\multicolumn{2}{|c|}{KoAT} &
\multicolumn{2}{|c|}{Rank} &
\multicolumn{2}{|c|}{Loopus} &
Speed &
AAA
\\
\hline

@@

@@

$1 + \max(x, y)$ &
$O(n)$ &

>>>>

We have experimentally evaluated the practicability of our automatic
amortized analysis with more than 30 challenging loop and recursion
patterns from open-source code and the
literature~\cite{GulwaniMC09,GulwaniJK09,GulwaniZ10}.  A full list of
examples together with the derived bounds is given in \pref{app:cat}.

\pref{fig:compar} shows four representative examples from the
evaluation.  Example \emph{t08} is a slightly modified version of
\emph{t08a} which is described in the introduction.  Example
\emph{t19} demonstrates the compositionaliy of the analysis.  The
program consists of two loops that decrement a variable $i$.  In the
first loop, $i$ is decremented down to 100 and in the second loop $i$
is increment further down to $-1$.  However, between the loops we
assign $i=i+k+50$.  So in total the program performs $52 + |[-1,i]| +
|[0,k]|$ increments.  Our analysis finds this tight bound because our
amortized analysis naturally takes into account the relation between
the two loops.  Other techniques derive a more conservative bound.

At first sight, example \emph{t30} appears to be a simple loop that
decrements the variable $x$ down to zero.  However, a closer look
reveals that the loop actually decrements both input variables $x$ and
$y$ down to zero before terminating.  In the loop body, first $x$ is
decremented by one.  Then the values of the variables $x$ and $y$ are
switched using the local variable $t$ as a buffer.  Our analysis
infers the tight bound $|[0,x]|+|[0,y]|$.  Sometimes we need some
assumptions on the inputs in order to derive a bound.  Example
\emph{t15} is such a case.  We assume here that the input variable $y$
is non-negative and write \code{assert(y>=0)}.  If we enter the loop
then we know that $x>0$ and we can obtain constant potential from the
assignment \code{x=x-1}.  After the assignment we know that $x\geq y$
and $y\geq 0$.  As a consequence, we can share the potential
$|[0,x]|$ before the assignment \code{x=x-y} between $|[0,x]|$ and
$|0,y|$ after the assignment.  In this way, we derive a tight linear
bound.

To compare our tool with existing work, we focused on loop bounds and
use a simple metric that counts the number of back edges (i.e., number
of loop iterations) that are follows in the execution of the program
because most other tools only bound this specific cost. In
\pref{fig:compar}, we show the bounds we derived (AAA) together with
the bounds derived by LOOPUS~\cite{SinnZV14} and
Rank~\cite{AliasDFG10}.  We also contacted the authors of SPEED but
have not been able to obtain this tool.
KoAT~\cite{BrockschmidtEFFG14} and PUBS~\cite{AlbertAGPZ12} currently
cannot operate on C code and the examples would need to be manually
translated into a term-rewriting to be analyzed by these tools.  For
Rank it is not totally clear how the computed bound relates to the C
program since the computed bound is for transitions in an automaton
that is derived from the C code.  For instance, the bound $2+y-x$ that
is derived for \emph{t08}.

\begin{figure*}
\centering
\begin{tabular}{r|c|c|c|c}
%
&
\begin{lstlisting}
while (y > x) {
  x++; tick(1); }
while (x > 2) {
  x -= 3; tick(1); }
\end{lstlisting}
&
\begin{lstlisting}
while (i > 100) {
  i--; tick(1) }
i += k+50;
while (i >= 0) {
  i--; tick(1); }
\end{lstlisting}
&
\begin{lstlisting}
while (x > 0) {
  x--;
  t=x, x=y, y=t;
  tick(1);
}
\end{lstlisting}
&
\begin{lstlisting}
assert(y >= 0);
while (x>y) {
  x -= y+1;
  for (z=y; z > 0; z--)
    tick(1);
  tick(1);
}
\end{lstlisting}
\\
%
\hline
& t08 (slightly modified t08a) & t19 & t30 & t15
%
\\
\hline
LOOPUS &
{\small $\max(0, x{-}2) {+} 2\max(0, y {-} x)$} &
{\small $\max(0, \! i{-}10^2) \!+ \max(0, \! k {+} i {+} 51)$} &
{\small % $1 + \max(x, y)$ (???)
---
} &
---
\\
%
%
%
%
Rank &
{\small $2 {+} y {-} x (?)$ } &
{\small $54+k+i$ } &
--- &
{\small $2+2x-y$ }
\\
%
%
AAA 
  & {\small $1.33|[x,y]| {+} 0.33 |[0,x]|$ }
  & $50 {+} |\inter {-1} i| {+} |\inter 0 k|$ 
  & $|\inter 0 x| {+} |\inter 0 y|$
  & $|\inter 0 x|$
\end{tabular}
\caption{ Comparison of resource bounds derived by different tools on
  several examples with linear bounds.  AAA stands for our automatic
  amortized analysis for Clight.  The output of Rank has been manually
  simplified to fit the table.  }
\label{fig:compar}
\end{figure*}

@@

\begin{figure}
\centering
\begin{tabular}{r|c|c|c|c|c}
& KoAT & Rank & LOOPUS & SPEED & AAA \\
\hline
\#bounds & 9 & 24 & 20 & 14 & 32
\\
\#lin.\ bounds & 9 & 21 & 20 & 14 & 32
\\
\#best bounds & 0 & 0 & 11 & 14 & 29
\\
\#tested & 14 & 33 & 33 & 14 & 33
\end{tabular}
\caption{Comparison of our automatic amortized analysis with other
  automatic tools.  We have not been able to obtain a version of
  SPEED~\cite{GulwaniMC09} and just use the bounds that have been
  reported by the authors. Similarly, KoAT~\cite{BrockschmidtEFFG14}
  does currently not work on C programs and we use the bounds that
  have been reported in the author's experimental evaluation.}
\label{fig:compar}
\end{figure}

@@

File &
\multicolumn{2}{|c|}{KoAT} &
\multicolumn{2}{|c|}{Rank} &
\multicolumn{2}{|c|}{LOOPUS} &
SPEED &
AAA
\\
\hline

@@

$\frac n m + m$ &

@@
